Generalized universal instability: Transient linear amplification and subcritical turbulence

### 2015-01-09 universalModeTurbulencePaper.tex
Added new figure 2 relating finite-shear and zero-shear growth rates.

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{ii}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\omegase}{\omega_{*}}
\newcommand{\nuee}{\nu_{ee}}
\newcommand{\energy}{\mathcal{E}}
\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\changed}[1]{#1}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\transpose}{\mathrm{T}}

@@

In the following sections, we establish three results concerning the drift wave system in slab geometry
with magnetic shear,
extending previous work by considering a kinetic description and exploring the role of transient linear amplification.
For these studies, the initial-value gyrokinetic code gs2 \cite{gs2} will be used
for numerical computations.
First, in section \ref{sec:transient} we demonstrate that
significant transient linear amplification is possible in this system,
as long as the magnetic shear is not too strong.
Second, in section \ref{sec:turbulence} we demonstrate that universal-mode turbulence without linear
instability, seen previously in the fluid model in \cite{Drake},
is also possible in the gyrokinetic model.
Finally, we prove there is a relationship between
transient linear amplification and nonlinear turbulence in this system,
using similar arguments to the proof for hydrodynamics in \cite{DelSoleNecessity}
and arriving at similar conclusions.
Specifically, if the physical parameters are such that turbulence can be sustained,
typical states of the turbulence must be amplified linearly for some amount of time, even if these states
eventually decay exponentially.
Two more precise statements of this result and their proof will be given in section \ref{sec:proof}.

@@

The system (\ref{eq:gke})-(\ref{eq:qn}) may or may not possess absolute linear instability,
depending on the physical parameters.
For the rest of this paper, we take $\eta_i = \eta_e = 0$
so it is certain that the ion-temperature-gradient (ITG) and electron-temperature-gradient (ETG)
instabilities are suppressed.
Even without temperature gradients, the system may still be unstable to the universal instability
driven by the density gradient.
For reference, the growth rates and frequencies of this instability
in the absence of magnetic shear are given in Appendix \ref{appendix:zeroShear}.
As this instability is stabilized for $k_y \rho_i \ll 1$ when even a small amount of magnetic shear
is introduced \cite{Ross, Tsang}, we might expect that this system can exhibit transient linear amplification, with
the number of $e$-foldings of amplification tending to infinity as the magnetic shear tends to zero.  In this way, the apparently singular limit
of zero shear may be seen as a continuous limit of the general shear case.

@@

@@

\begin{figure}[h!]
% Data computed on Edison in /global/scratch2/sd/landrema/20141121-01-coderunner-edison-transientAmplificationScans
% Plot generated by first running
% m20141121_02_plotTransientAmplificationLsOverLnScanForPaper.m on edison,
% and then running
% m20140723_03_plotTransientAmplificationLsOverLnScanForPaper.m on my laptop.
% ----------------------------------------
% Data for previous version of figure computed on Edison in /global/scratch2/sd/landrema/20140718-01-coderunner-edison-transientAmplificationScans
% Plot generated by first running
% m20140725_02_plotTransientAmplificationLsOverLnScanForPaper.m on edison,
% and then running
% m20140723_03_plotTransientAmplificationLsOverLnScanForPaper.m on my laptop.
\includegraphics[width=3in]{m20141121_03_plotTransientAmplificationLsOverLnScanForPaper_a.pdf}
\includegraphics[width=3in]{m20141121_03_plotTransientAmplificationLsOverLnScanForPaper_b.pdf}
%\includegraphics[width=3in]{m20140723_03_plotTransientAmplificationLsOverLnScanForPaper.pdf}
\caption{(Color online) The (a) number of $e$-foldings of transient amplification and (b)
duration of the amplification both scale with the magnetic shear length scale $L_s$,
consistent with the appearance of absolute instability in the limit of vanishing shear ($L_s \to \infty$).
Parameters used are $k_y \rho_i = 2$ and $\nuee = 0.05 v_i / L_n$.
\label{fig:transientAmplificationVsLsOverLn}}
\end{figure}

@@

The ``universal mode'' system (slab geometry with a density gradient, electrostatic,
no sheared equilibrium flow, and weak temperature gradients)
is known to have a dramatic change in linearly stability
as the magnetic shear is raised from zero to small finite values
\cite{Ross, Tsang, Antonsen,usUniversalInstability}.
As the eigenvalues are so sensitive to magnetic shear,
we might expect traditional modal analysis to
give a misleading picture of typical behavior,
as with other systems that have sensitive eigenvalues \cite{TrefethenEmbree}.
Indeed, here we have demonstrated that significant
transient linear amplification is possible in the drift wave system even when
all eigenmodes are decaying.
We have demonstrated that the amount and duration of amplification scale
with the magnetic shear length $L_s$, so these quantities naturally become
unbounded in the limit of vanishing shear.
The non-modal amplification also decreases with increasing collisionality,
particularly at the shortest perpendicular wavelengths,
and the amplification tends to be largest at wavelengths near $k_y \rho_i \sim 1-2$.
In contrast to more famous examples of non-modal amplification
from hydrodynamics \cite{TrefethenSubcritical}, and to some recent examples from plasma physics \cite{Newton, Alex, BarnesRotation, EdmundPRL},
the amplification in the plasma drift wave system here
does not depend on sheared equilibrium flow.

@@

\begin{acknowledgments}
This material is based upon work supported by the
U.S. Department of Energy, Office of Science, Office of Fusion Energy Science,
under Award Numbers DEFG0293ER54197 and DEFC0208ER54964.
Computations were performed on the Edison system at
the National Energy Research Scientific Computing Center, a DOE Office of Science User Facility supported by the Office of Science of the U.S. Department of Energy under Contract No. DE-AC02-05CH11231.
Conversations with 
Tom Antonsen, Michael Barnes, Greg Colyer, James Drake, Edmund Highcock, Greg Hammett, Adil Hassam, and Freja Nordsiek
are gratefully acknowledged.
\end{acknowledgments}

@@

>>>>

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{ii}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\omegase}{\omega_{*}}
\newcommand{\nuee}{\nu_{ee}}
\newcommand{\energy}{\mathcal{E}}
\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\changed}[1]{\textcolor{red}{#1}}
\newcommand{\transpose}{\mathrm{T}}

@@

In the following sections, we establish three results concerning the drift wave system in slab geometry
with magnetic shear,
extending previous work by considering a kinetic description and exploring the role of transient linear amplification.
For these studies, the initial-value gyrokinetic code gs2 \cite{gs2} will be used
for numerical computations.
First, in section \ref{sec:transient} we demonstrate that
significant transient linear amplification is possible in this system,
\changed{and that the transiently growing fluctuations
resemble the zero-shear eigenmodes
in their dependence of the growth rate on wavenumber.}
%as long as the magnetic shear is not too strong.
Second, in section \ref{sec:turbulence} we demonstrate that universal-mode turbulence without linear
instability, seen previously in the fluid model in \cite{Drake},
is also possible in the gyrokinetic model.
Finally, we prove there is a relationship between
transient linear amplification and nonlinear turbulence in this system,
using similar arguments to the proof for hydrodynamics in \cite{DelSoleNecessity}
and arriving at similar conclusions.
Specifically, if the physical parameters are such that turbulence can be sustained,
typical states of the turbulence must be amplified linearly for some amount of time, even if these states
eventually decay exponentially.
Two more precise statements of this result and their proof will be given in section \ref{sec:proof}.

@@

The system (\ref{eq:gke})-(\ref{eq:qn}) may or may not possess absolute linear instability,
depending on the physical parameters.
For the rest of this paper, we take $\eta_i = \eta_e = 0$
so it is certain that the ion-temperature-gradient (ITG) and electron-temperature-gradient (ETG)
instabilities are suppressed.
Even without temperature gradients, the system may still be unstable to the universal instability
driven by the density gradient.
For reference, the growth rates and frequencies of this instability
in the absence of magnetic shear are given in Appendix \ref{appendix:zeroShear}.
\changed{
As detailed in the appendix, some insight into the instability can be obtained by expanding
the the plasma dispersion function $Z$ for the electrons in the small-argument limit,
$Z(\omega / (|k_{||}| v_e)) \approx i \sqrt{\pi}$ for $|\omega / (k_{||} v_e)| \ll 1$,
and by expanding the ion $Z$ function in the large argument limit,
$Z(\omega / (|k_{||}| v_i)) \approx - |k_{||}| v_i / \omega$ for $|\omega / (k_{||} v_i)| \gg 1$.
One then finds a frequency with positive imaginary part $\propto i \sqrt{\pi} \omega_*$,
indicating the instability is due to resonant electrons in the presence of a density gradient.}
As this instability is stabilized for $k_y \rho_i \ll 1$ when even a small amount of magnetic shear
is introduced \cite{Ross, Tsang}, we might expect that this system can exhibit transient linear amplification, with
the number of $e$-foldings of amplification tending to infinity as the magnetic shear tends to zero.  In this way, the apparently singular limit
of zero shear may be seen as a continuous limit of the general shear case.

@@

\changed{
The transiently growing structures in the presence of magnetic shear
are closely related to the zero-shear eigenmodes.  To demonstrate this relationship, we define an effective
$k_{\perp}$ for the transient fluctuation at each time by
\begin{equation}
\label{eq:expectedKperp}
\left< k_{\perp} \right>(t) = \frac{\int dz\; k_{\perp}(z) \left| \Phi(z,t)\right|^2}
{\int dz\; k_{\perp}(z) \left| \Phi(z,t)\right|^2}.
\end{equation}
Similarly, we define an effective $|k_{||}|$ at each time for the transient structure
by
\begin{equation}
\label{eq:expectedKpar}
\left< \left| k_{||} \right| \right>(t) = \frac{\int dk_{||}  \left| k_{||}\right| \left| \tilde\Phi(k_{||},t)\right|^2}
{\int dk_{||}\left| \tilde\Phi(k_{||},t)\right|^2}
\end{equation}
where $\tilde\Phi$ is the Fourier transform of $\Phi$ in $z$. Using these effective values
of $k_{\perp}$ and $k_{||}$, the zero-shear dispersion relation may be evaluated.
The resulting growth rate can then be compared to the instantaneous growth rate of the finite-shear
transient structure, $\gamma = (2W(t))^{-1} dW/dt$.  This comparison is shown in figure \ref{fig:comparingTransientTo0Shear}.
In this figure, we display 25 independent simulations of transients,
varying $L_s/L_n$ between 20 and 300, and plotting the evolution of $\gamma$ from the time
of peak growth rate to the time at which the energy begins to decay.
In each case, there is very close agreement between the instantaneous growth rate of the transient
and the `predicted' growth rate from the zero-shear dispersion relation.
Thus, we can conclude that the transiently growing structures are closely related to the zero-shear
eigenmodes.  Magnetic shear causes $k_{\perp}$ and $k_{||}$ to vary with time as the structure
evolves (unlike the zero-shear case in which $k_{\perp}$ and $k_{||}$ are constants.)
However, at each instant, the finite-shear transient grows at precisely the rate expected from the zero-shear
instability at the average wavenumbers $\left< k_{\perp} \right>$ and $\left< \left| k_{||} \right| \right>$.
For the simulations in figure  \ref{fig:comparingTransientTo0Shear}, $k_y \rho_i = 2$ and $\nuee = 0.05 v_i/L_n$.
}

\begin{figure}[h!]
% Data generated on Edison in /global/scratch2/sd/landrema/20150109-01-coderunner-edison-0shearUniversalInstability/vnewk_e_0p05
% Figure generated with m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.m
\includegraphics[width=3in]{m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.pdf}
\caption{(Color online)
\changed{
Although a change from zero to small finite magnetic shear causes a fundamental change in the eigenvalues,
(which change from unstable to decaying), 
the \emph{transient} finite-shear structures remain similar to the zero-shear modes.  For as shown here,
the instantaneous growth rate of the transient finite-shear structure is nearly identical
to the growth rate of a zero-shear eigenmode at the dominant $k_{\perp}$ and $k_{||}$ values
(\ref{eq:expectedKperp})-(\ref{eq:expectedKpar}).  Each trajectory shown is an independent
linear initial-value calculation, with $L_s/L_n$ varying from 20-300 between simulations.  The growth rate is plotted from
the time of maximum growth rate (at the $+$ symbols) until decay begins.
}
\label{fig:comparingTransientTo0Shear}}
\end{figure}

\changed{
For the range of $k_{||}$ and $k_{\perp}$ values computed for the transient structures
of figure \ref{fig:comparingTransientTo0Shear}, the zero-shear dispersion relation
is well approximated by the approximate analytic dispersion relation (\ref{eq:approxRoot}).  Thus, we can also conclude
that the transient structures, like the zero-shear eigenmodes,
are driven by the combination of resonant electrons and the density gradient.
}

@@

\begin{figure}[h!]
% Data computed on Edison in /global/scratch2/sd/landrema/20141121-01-coderunner-edison-transientAmplificationScans
% Plot generated by first running
% m20141121_02_plotTransientAmplificationLsOverLnScanForPaper.m on edison,
% and then running
% m20141121_03_plotTransientAmplificationLsOverLnScanForPaper.m on my laptop.
% ----------------------------------------
% Data for previous version of figure computed on Edison in /global/scratch2/sd/landrema/20140718-01-coderunner-edison-transientAmplificationScans
% Plot generated by first running
% m20140725_02_plotTransientAmplificationLsOverLnScanForPaper.m on edison,
% and then running
% m20140723_03_plotTransientAmplificationLsOverLnScanForPaper.m on my laptop.
\includegraphics[width=3in]{m20141121_03_plotTransientAmplificationLsOverLnScanForPaper_a.pdf}
\includegraphics[width=3in]{m20141121_03_plotTransientAmplificationLsOverLnScanForPaper_b.pdf}
%\includegraphics[width=3in]{m20140723_03_plotTransientAmplificationLsOverLnScanForPaper.pdf}
\caption{(Color online) The (a) number of $e$-foldings of transient amplification and (b)
duration of the amplification both scale with the magnetic shear length scale $L_s$,
consistent with the appearance of absolute instability in the limit of vanishing shear ($L_s \to \infty$).
Parameters used are $k_y \rho_i = 2$ and $\nuee = 0.05 v_i / L_n$.
\label{fig:transientAmplificationVsLsOverLn}}
\end{figure}

@@

The ``universal mode'' system (slab geometry with a density gradient, electrostatic,
no sheared equilibrium flow, and weak temperature gradients)
is known to have a dramatic change in linearly stability
as the magnetic shear is raised from zero to small finite values
\cite{Ross, Tsang, Antonsen,usUniversalInstability}.
As the eigenvalues are so sensitive to magnetic shear,
we might expect traditional modal analysis to
give a misleading picture of typical behavior,
as with other systems that have sensitive eigenvalues \cite{TrefethenEmbree}.
Indeed, here we have demonstrated that significant
transient linear amplification is possible in the drift wave system even when
all eigenmodes are decaying.
\changed{
As shown in figure \ref{fig:comparingTransientTo0Shear},
the transient structures grow at the zero-shear
growth rate, when evaluated at the dominant $k_{\perp}$ and $k_{||}$ at that instant.
}
We have demonstrated that the amount and duration of amplification scale
with the magnetic shear length $L_s$, so these quantities naturally become
unbounded in the limit of vanishing shear.
The non-modal amplification also decreases with increasing collisionality,
particularly at the shortest perpendicular wavelengths,
and the amplification tends to be largest at wavelengths near $k_y \rho_i \sim 1-2$.
In contrast to more famous examples of non-modal amplification
from hydrodynamics \cite{TrefethenSubcritical}, and to some recent examples from plasma physics \cite{Newton, Alex, BarnesRotation, EdmundPRL},
the amplification in the plasma drift wave system here
does not depend on sheared equilibrium flow.

@@

\begin{acknowledgments}
This material is based upon work supported by the
U.S. Department of Energy, Office of Science, Office of Fusion Energy Science,
under Award Numbers DEFG0293ER54197 and DEFC0208ER54964.
Computations were performed on the Edison system at
the National Energy Research Scientific Computing Center, a DOE Office of Science User Facility supported by the Office of Science of the U.S. Department of Energy under Contract No. DE-AC02-05CH11231.
Conversations with 
Tom Antonsen, Michael Barnes, Greg Colyer, James Drake, Edmund Highcock, Greg Hammett, and Adil Hassam
are gratefully acknowledged.
\end{acknowledgments}

@@

\changed{
Some insight into the density-gradient-driven instability can be obtained by
considering the limit $|\zeta_e| \ll 1$ so $Z_e \approx i\sqrt{\pi}$,
and $|\zeta_i| \gg 1$ so $Z_i \approx -1/\zeta_i$. Considering solutions with $|\omega/\omega_*| \ll 1$,
one finds
\begin{equation}
\label{eq:approxRoot}
\omega = \frac{\omega_* \Gamma_{0i}}
{\left(\frac{T_e}{T_i}+1\right)^2 + \left(\sqrt{pi}\frac{\omega_*}{|k_{||}| v_e}\right)^2}
\left( \frac{T_e}{T_i} + 1 + i \sqrt{\pi}\frac{\omega_*}{|k_{||}| v_e}\right).
\end{equation}
The imaginary part of $\omega$ evidently arises due to the electron resonance $Z_e \approx i\sqrt{\pi}$.
}
